Cultural Evolution of Conservation Behaviors Within and Between Communities

### 2021-04-12 DissertationProposal/DissertationProposal.Rmd
cleaning up

Researchers have posited that cultural group selection theory should be applicable to understanding the spread and persistence of conservation behaviors such as adherence to community conservation areas or adoption of new efficient technologies [@kline2018designing]. Cultural group selection theory offers mechanistic insights for the emergence and spread of behaviors within and between groups broadly, but its applicability to conservation specifically is unknown. Conservation has many aspects which make it unique compared to behavioral adoption in the abstract. In conservation, an individual’s behavior directly affects their environment, which may in turn become altered in a way which shapes future individual decision making. Feedbacks between individual behavior and resource availability are further complicated by resource characteristics such as carrying capacity, mobility, and regeneration rate. These specifics make conservation a fundamentally different system from general behavioral adoption and cooperation, thus we must extend cultural group selection theory considering conservation explicitly. The aim of this research is to expand the theories of behavioral adoption via cultural group selection to explicitly consider factors unique to conservation, and to empirically examine the mechanisms of conservation behavioral adoption within and between groups. This research will offer insights for strategically seeding both community conservation areas and new technologies so that they may offer the greatest benefit to communities and the preservation of renewable resources.

@@

```{r moon, echo=FALSE, fig.cap="Pemba Island, Tanzania. Showing wards with and without CoFMAs. For each CoFMA, the size (ha) is proportional to the size of the circle and the percent area designated as Use and Conservation is indicated by the blue and green colors respectively.", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/MoonMapEditedNoLabs.png")
```

@@

# Chapter 1: Theoretical Advances for Applying Cultural Group Selection Theory to Community Conservation Areas. 
## Background
For community conservation areas to tangibly reduce the loss of earth’s natural resources, they must be adopted at large spatial scales and remain intact in perpetuity [@turner2007global]. Predicting if and how conservation actions will spread and evolve across a landscape has proven to be a major challenge for conservation scientists and practitioners [@child2010conceptual]. Theory around the long-term adoption of conservation norms is sparse [@mahajan2020theory]. Because of this gap, most researchers default to the diffusion of innovations literature to understand and draw conclusions around conservation adoption [@mascia2018conservation; @rogers_diffusion_2010]. This body of literature has proven valuable for demonstrating *how* conservation paradigms spread, but is relatively mute for understanding the underlying social mechanisms of *why* conservation paradigmes spread and last [@pannell2006understanding; @mills2019conservation].

@@

```{r FirstABMb, out.width = "100%",echo=FALSE, fig.cap="Resource quantity and mean harvest vales over time for one run of the agent-based model at a maximum harvest valuue one above the calculated value for sustainable harvest. Model starting conditions include 100 individuals harvesting from a landscape of 10000 resources. The regeneration factor at each time step is 1.20.", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/abmP1.png")
```

@@

Model dynamics show that over time, agents begin to harvest from the conservation area as they copy more successful individuals. This behavior depletes the resource pool in the conservation area and lowers the number of resources which regenerate and agents can harvest in the subsequent time step. Note that full defection is defined as agents spending an equal proportion of their effort in the conservation area as the proportion conservation (i.e. they treat the entire landscape equally).

@@

In the competing groups scenario, the model will follow the exact same scheduling process as in the separate groups scenario, but a specified portion of the working landscape(s) will be shared among the groups. Mobile resources leaving the conservation areas will emigrate to the working landscape as in the single group model. If excess resources exist in the working landscape, they will move to whichever conservation area is below its carrying capacity. If both conservation areas are below carrying capacity, resources will have an equal chance of moving to either. Lastly, the carrying capacity of the working landscape associated with each group will include the shared area, but not the other group’s working landscape. This will ensure that resources do not always reach an equilibrium between the groups and the groups may have variable success. See figure \@ref(fig:conceptDiagram2) for a conceptual description of the two multi-group scenarios.

@@

# Chapter 2: Effects of Inter-Group Resource Competition and Selection on Community Adoption of Costly Conservation Policy.

## Background
Human social structures are characterized by nested groups, ranging from a nuclear family unit to a national or even multi-national identity. Group behavioral norms can have profound impacts on the wellbeing of group members and their environment, far beyond that of a single individual. Cultural group selection theorizes, and to a small extent has empirically demonstrated, the mechanisms which establish and perpetuate group-level norms [@henrich_cultural_2004; @bell_culture_2009]. Most relevant to community conservation is the development of norms which are beneficial for the group, but cause an individual to incur some personal cost; we refer to these as costly cooperative behaviors [@molleman_cultural_2013]. Central to theories around the adoption of costly cooperative behaviors is the insight that selection acts on both individuals and groups. Individuals may engage in personally costly behavior if it is beneficial for their group, which they also depend on for their wellbeing [@handley_human_2020]. The degree to which an individual depends on each of their nested groups is directly proportional to the personal costs they will incur for that group [@okasha_multilevel_2004; @price1972extension]. To the extent that conservation is a collaborative dilemma, these insights from cultural group selection may be paramount to implementing effective community based conservation [@waring2015multilevel; @kline2018designing].

Conservation is a particularly wicked collection action problem because the group level costs and benefits operate at a much longer time scale than those of the individual. The group level benefits of conservation are often much less certain than the immediate individual costs of forgoing resources. Humans tend to steeply discount the benefits of their actions as those benefits move further away in time and the benefits become more uncertain [@dasgupta2005uncertainty]. This hyperbolic discounting makes conservation norms very unlikely to emerge and persist under our current understanding of cultural group selection — the individual costs are high and group benefits are low. The role of institutions in human society is to reduce these barriers and to encourage collaborative behavior [@north1990institutions; @ostrom1990governing]. One particular institution aimed at promoting costly cooperative behaviors around conservation is REDD+ [@angelsen2017redd; @angelsen2008moving].

@@

Leakage [as defined by @andam2008measuring] is a form of land use spillover where an intervention, such as the CoFMAs established on Pemba by REDD+, has a negative, counteracting effect on the targeted outcome variable in another place. On Pemba, we have anecdotal testimonies that individuals in wards with CoFMAs have simply shifted their harvest pressure onto adjacent wards. While dubious for the efficacy of REDD+ writ large, on Pemba, leakage may actually be driving the spread of community based conservation observed in the ten new communities which elected to create CoFMAs post-REDD+ [@andrewsdoes]. Just as the performance based payments from REDD+ were designed to alter the individual/community cost/benefit matrix so that selection favors costly collaboration via immediate community payments, we hypothesize that leakage from wards with CoFMAs may also act to favor costly cooperative behavior as individuals must act to defend their access to resources they would not have otherwise.

@@

To better assess if individuals are adhering to CoFMA regulations, we need to sort out the originally demarcated use and conservation areas in each CoFMA. To our knowledge, these data do not currently exist or are not available. As a first step in this research, we will attempt to access this information through the Department of Forestry and Non-Renewable Natural Resources libraries on Pemba. Where/if this data is not available, we will drive or walk the boundary of each area to create a shapefile to use for analysis.

@@

# Chapter 3: Social Influences on Transmission Fidelity in Household Level Diffusion of Beneficial Technology.

@@

While transmission fidelity can vary significantly given the context, it is often unclear what effect this variation has on the outcomes of imitated behavior. Theoretical models have shown that even small changes in transmission fidelity can have strong impacts on population-level behavioral outcomes [@lewis_transmission_2012]. Alternatively, some behaviors require only very low-fidelity replication to achieve identical outcomes [@enquist_one_2010]. In fact, it is likely that if a behavior is under strong environmental constraints then individual learning should help correct errors in the transmission process.

@@

The survey instrument used for this research will elicit information regarding the relationship between the builder of each cookstove and the individual they copied their stove design from. Specifically, this survey tool will be used to gather data on geographic distance and communication frequency. Geographic distance will be measured according to the shortest walking distance between the two stoves. Communication frequency will be measured in terms of weekly communication and taken as a measure of social connectedness between the original stove builder and the copying individual. I will also use this survey to collect data on the decision making processes individuals exhibit when deciding to build a fuel-efficient cookstove.

@@

The key research questions for this project will be assessed using multilevel regression operationalized in a Bayesian context. I will compare standardized parameter estimates, to determine which factors (geographic distance or social relatedness) most strongly contribute to copying fidelity in this study system. Similar analyses will be used to determine which factors, if any, are strongly associated with stove fuel efficiency.

>>>>

Researchers have posited that cultural group selection theory should be applicable to understanding the spread and persistence of conservation behaviors such as adherence to community conservation areas or adoption of new efficient technologies [@kline2018designing]. Cultural group selection theory offers mechanistic insights for the emergence and spread of behaviors within and between groups broadly, but its applicability to conservation is unknown. Conservation has many aspects which make it unique compared to behavioral adoption in the abstract. In conservation, an individual’s behavior affects their environment, which may in turn become altered in a way which shapes future individual decision making. Feedbacks between individual behavior and resource availability are further complicated by resource characteristics such as carrying capacity, mobility, and regeneration rate. These specifics make conservation a fundamentally different system from general behavioral adoption and cooperation, thus we must extend cultural group selection theory considering conservation. The aim of this research is to expand the theories of behavioral adoption via cultural group selection to explicitly consider factors unique to conservation, and to empirically examine the mechanisms of conservation behavioral adoption within and between groups. This research will offer insights for implementing both community conservation areas and new technologies so that they may offer the greatest benefit to communities and the preservation of renewable resources.

@@

```{r moon, echo=FALSE, fig.cap="Pemba Island, Tanzania. Showing wards with and without CoFMAs. For each CoFMA, the size (ha) is proportional to the size of the circle and the percent area designated as \\textit{use} and \\textit{conservation} is indicated by the blue and green colors respectively.", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/MoonMapEditedNoLabs.png")
```

@@

# Chapter 1: Theoretical Advances for Applying Cultural Group Selection Theory to Community Conservation Areas 
## Background
For community conservation areas to tangibly reduce the loss of earth’s natural resources, they must be adopted at large spatial scales and remain intact in perpetuity [@turner2007global]. Predicting if and how conservation actions will spread and evolve across a landscape has proven to be a major challenge for conservation scientists and practitioners [@child2010conceptual]. Theory around the long-term adoption of conservation norms is sparse [@mahajan2020theory]. Because of this gap, most researchers default to the diffusion of innovations literature to understand and draw conclusions around conservation adoption [@mascia2018conservation; @rogers_diffusion_2010]. This body of literature has proven valuable for demonstrating *how* conservation paradigms spread, but is relatively mute for understanding the underlying social mechanisms of *why* conservation paradigmes spread and last [@pannell2006understanding; @mills2019conservation].

@@

```{r FirstABMb, out.width = "100%",echo=FALSE, fig.cap="Resource quantity and mean harvest vales over time for one run of the agent-based model at a maximum harvest value one above the calculated value for sustainable harvest. Model starting conditions include 100 individuals harvesting from a landscape of 10000 resources. The regeneration factor at each time step is 1.20.", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/abmP1.png")
```

@@

Model dynamics show that over time, agents begin to harvest from the conservation area as they copy more successful individuals. This behavior depletes the resource pool in the conservation area and lowers the number of resources which regenerate and agents can harvest in the subsequent time step. Note that full defection is when agents spending an equal proportion of their effort in the conservation area as the proportion of resources conserved (i.e. they treat the entire landscape equally).

@@

In the competing groups scenario, the model will follow the same scheduling process as in the separate groups scenario, but a specified portion of the working landscape(s) will be shared among the groups. Mobile resources leaving the conservation areas will emigrate to the working landscape as in the single group model. If excess resources exist in the working landscape, they will move to whichever conservation area is below its carrying capacity. If both conservation areas are below carrying capacity, resources will have an equal chance of moving to either. Lastly, the carrying capacity of the working landscape associated with each group will include the shared area, but not the other group’s working landscape. This will ensure that resources do not always reach an equilibrium between the groups and the groups may have variable success. See figure \@ref(fig:conceptDiagram2) for a conceptual description of the two multi-group scenarios.

@@

# Chapter 2: Effects of Inter-Group Resource Competition and Selection on Community Adoption of Costly Conservation Policy

## Background
Human social structures are characterized by nested groups, ranging from a nuclear family unit to a national or even multi-national identity. Group behavioral norms can have profound impacts on the wellbeing of group members and their environment, far beyond that of a single individual. Cultural group selection theorizes, and to a small extent has empirically demonstrated, the mechanisms which establish and perpetuate group-level norms [@henrich_cultural_2004; @bell_culture_2009]. Most relevant to community conservation is the development of norms which are beneficial for the group, but cause individuals to incur some personal cost; we refer to these as costly cooperative behaviors [@molleman_cultural_2013]. Central to theories around the adoption of costly cooperative behaviors is the insight that selection acts on both individuals and groups. Individuals may engage in personally costly behavior if it is beneficial for their group, which they also depend on for their wellbeing [@handley_human_2020]. The degree to which an individual depends on each of their nested groups is directly proportional to the personal costs they will incur for that group [@okasha_multilevel_2004; @price1972extension]. To the extent that conservation is a collaborative dilemma, these insights from cultural group selection may be paramount to implementing effective community-based conservation [@waring2015multilevel; @kline2018designing].

Conservation is a particularly wicked collection action problem because the group level costs and benefits operate at a much longer time scale than those of the individual. The group level benefits of conservation are often much less certain than the immediate individual costs of forgoing resources. Humans tend to steeply discount the benefits of their actions as those benefits move further away in time and the benefits become less certain [@dasgupta2005uncertainty]. This hyperbolic discounting makes conservation norms very unlikely to emerge and persist under our current understanding of cultural group selection — the individual costs are high and group benefits are low. The role of institutions in human society is to reduce these barriers and to encourage collaborative behavior [@north1990institutions; @ostrom1990governing]. One particular institution aimed at promoting costly cooperative behaviors around conservation is REDD+ [@angelsen2017redd; @angelsen2008moving].

@@

Leakage [as defined by @meyfroidt2020focus] is a form of land use spillover where an intervention, such as the CoFMAs established on Pemba by REDD+, has a negative, counteracting effect on the targeted outcome variable in another place. On Pemba, we have anecdotal testimonies that individuals in wards with CoFMAs have simply shifted their harvest pressure onto adjacent wards. While dubious for the efficacy of REDD+ writ large, on Pemba, leakage may actually be driving the spread of community-based conservation observed in the ten new communities which elected to create CoFMAs post-REDD+ [@andrewsdoes]. Just as the performance based payments from REDD+ were designed to alter the individual/community cost/benefit matrix so that selection favors costly collaboration via immediate community payments, we hypothesize that leakage from wards with CoFMAs may also act to favor costly cooperative behavior as individuals must act to defend their access to resources they would not have otherwise.

@@

To better assess if individuals are adhering to CoFMA regulations, we need to map the originally demarcated use and conservation areas in each CoFMA. To our knowledge, these data do not currently exist or are not available. As a first step in this research, we will attempt to access this information through the Department of Forestry and Non-Renewable Natural Resources libraries on Pemba. Where/if this data is not available, we will drive or walk the boundary of each area to create a shapefile to use for analysis.

@@

# Chapter 3: Social Influences on Transmission Fidelity in Household Level Diffusion of Beneficial Technology

@@

While transmission fidelity can vary significantly given the context, it is often unclear what effect this variation has on the outcomes of imitated behavior. Theoretical models have shown that even small changes in transmission fidelity can have strong impacts on population-level behavioral outcomes [@lewis_transmission_2012]. Alternatively, some behaviors require only very low-fidelity replication to achieve identical outcomes [@enquist_one_2010]. In fact, it is likely that if a behavior is under strong environmental constraints then individual learning should help correct errors in the transmission process; therefore, functional traits which directly affect an individual’s wellbeing should be more resilient to modification during transmission [@andersson2013fidelity].

@@

The survey instrument used for this research will elicit information regarding the relationship between the builder of each cookstove and the individual they copied their stove design from. Specifically, this survey tool will be used to gather data on geographic distance and communication frequency. Geographic distance will be measured according to the shortest walking distance between the two stoves. Communication frequency will be measured in terms of weekly communication of any type and taken as a measure of social connectedness between the original stove builder and the copying individual. As recall data tends to be unreliable, I will use current social connectedness as opposed to the time period the stove was built. I will also use this survey to collect data on the decision making processes individuals exhibit when deciding to build a fuel-efficient cookstove. Lastly, a variety of external factors may contribute to the outcome of cookstove construction. I will collect information to control for household wealth, access to materials, and intended cookstove use.

@@

The key research questions for this project will be assessed using multilevel regression operationalized in a Bayesian context. I will compare standardized parameter estimates, to determine which factors (geographic distance or social relatedness) most strongly contribute to copying fidelity in this study system. Similar analyses will be used to determine if these same factors, if any, are strongly associated with stove fuel efficiency.
